Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2018-09-19 appendix.tex
version 3...

For the $i$th factor of \eqref{eq:full_likelihood}, we decompose the likelihood into its marginal and conditional components as
\[
\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
\]
As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
\begin{align*}
    \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t)
\end{align*}
where $\overline{Y}$ is the complement of the site pattern--valued random variable $Y$ and has the same distribution as $Y$.
Note
\begin{align*}
    \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Thus given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subseteq\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, i.e., $\ancestralSplitPartition_j(\tau, t)$ is a set of sets of most likely internal node labels.
Here, $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,\nSiteRows}=0$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
The same can be done for the case of $\alignmentColumn_{i,\nSiteRows}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective.
From this we have
\begin{align*}
    \max_{\ancestralStateColumn_i} \ \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t)\\
    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)\\
    &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)
\end{align*}
for some $j$.
Thus, each term in the likelihood can be collapsed into terms relating only to site splits and ancestral state splits, indexed by $j$, as opposed to individual observations, indexed by $i$.

\subsection*{Example}
We follow with an example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
\[
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
\]
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case depends on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
\]
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral state splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
The probabilities of observing ancestral state splits for $\siteSplit_1=\emptyset$ are
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
\]
\[
\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
\]
\[
\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
\]
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
\[
\Pr(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
\]
For $\siteSplit_5=\{1,2\}$ we have
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3,
\]
\[
\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^3(1-p)^2.
\]
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
\[
\Pr(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3.
\]

@@

The likelihood in \eqref{eq:profile_likelihood} can be written as
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
for $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ and some $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

Let
\[
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
\]
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the marginal log-likelihood
\[
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
\]
and the conditional log-likelihood
\[
\ell^C_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
\]
with $\boldsymbol\xi = [\xi_j]_{j=1}^\nSiteSplits$ so that \eqref{eq:site_pattern_profile_likelihood_mean} is
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^C_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\end{equation}

@@

To compute the likelihood of observing a set of data, we need $\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
Using branch fidelities, the probability of a character change along a branch with fidelity parameter $x$ is $(1-x)/2$, while the probability of a character remaining the same is $(1+x)/2$.
See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
Conditional likelihood computations for all site splits and ancestral state splits are in Table~\ref{tab:farris_likelihoods} for the InvFels tree.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like00}
\caption[short]{}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like10}
\caption[short]{}
\end{subfigure}
\caption{
    Example likelihood computations on the InvFels tree $\tau_1$ for fidelities $t=\{x_1,y_1,x_2,y_2,w\}$.
    Edges labeled by the probability of substitution along that edge.
    In (a), we compute the product to obtain $\Pr(\ancestralSplitRV=\emptyset\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x_1)(1-x_2)(1+y_1)(1-y_2)(1+w)/32$.
    In (b), the same process yields $\Pr(\ancestralSplitRV=\{1\}\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x_1)(1-x_2)(1+y_1)(1-y_2)(1-w)/32$.
}
\label{fig:example_likelihoods}
\end{figure}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\hline
$\siteSplit_j$ & $\ancestralSplit_k$ & $32\cdot\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_1, t)$\\
\hline
$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
&$\{1\}^*$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$             \\
&$\{2\}^*$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$           \\

@@

We first prove intermediate results to show the main result.

\begin{lemma}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
There exists a set of $0 < x^*, y^* < 1$ such that $\emptyset$ is the maximal ancestral state split for all site splits given $\tau=\tau_1$.
\label{lemma:ancestral-state}
\end{lemma}

>>>>

We now consider the $i$th factor of \eqref{eq:full_likelihood}.
As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
\begin{align*}
    \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)\\
    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)
\end{align*}
where $\overline{Y}$ is the complement of the site pattern--valued random variable $Y$ and has the same distribution as $Y$.
Since
\begin{align*}
    \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t),
\end{align*}
given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_j \mid \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subseteq\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, i.e., $\ancestralSplitPartition_j(\tau, t)$ is a set of sets of most likely internal node labels.
Here, $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,\nSiteRows}=0$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
The same can be done for the case of $\alignmentColumn_{i,\nSiteRows}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective.
From this we have
\begin{align*}
    \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t)\\
    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)\\
    &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_j \mid \tau, t)
\end{align*}
for some $j$.
Thus, each term in the likelihood can be collapsed into terms relating only to site splits and ancestral state splits, indexed by $j$, as opposed to individual observations, indexed by $i$.

\subsection*{Example}
We follow with an example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
\[
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
\]
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case depends on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
\[
\Pr(\siteSplitRV=\siteSplit, \ancestralSplitRV=\emptyset \mid \tau, t)\neq \Pr(\siteSplitRV=\siteSplit, \ancestralSplitRV=\{1,2\} \mid \tau, t)
\]
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral state splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
The probabilities of observing ancestral state splits for $\siteSplit_1=\emptyset$ are
\[
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\emptyset \mid \tau, t) =
(1-p)^5,
\]
\[
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\{1\} \mid \tau, t) =
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\{2\} \mid \tau, t) =
p^3(1-p)^2,
\]
\[
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\{1,2\} \mid \tau, t) =
p^4(1-p).
\]
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
\[
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\xi_1 \mid \tau, t) =
\Pr(\siteSplitRV=\emptyset, \ancestralSplitRV=\emptyset \mid \tau, t) =
(1-p)^5.
\]
For $\siteSplit_5=\{1,2\}$ we have
\[
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\emptyset \mid \tau, t) =
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\{1,2\} \mid \tau, t) =
p^2(1-p)^3,
\]
\[
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\{1\} \mid \tau, t) =
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\{2\} \mid \tau, t) =
p^3(1-p)^2.
\]
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
\[
\Pr(\siteSplitRV=\{1,2\}, \ancestralSplitRV=\xi_5 \mid \tau, t) =
p^2(1-p)^3.
\]

@@

The likelihood in \eqref{eq:profile_likelihood} can be written as
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
for $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ and some $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

Let
\[
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)}
\]
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}

@@

To compute the likelihood of observing a set of data, we need $\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
Using branch fidelities, the probability of a character change along a branch with fidelity parameter $x$ is $(1-x)/2$, while the probability of a character remaining the same is $(1+x)/2$.
See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
Likelihood computations for all site splits and ancestral state splits are in Table~\ref{tab:farris_likelihoods} for the InvFels tree.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like00}
\caption[short]{}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like10}
\caption[short]{}
\end{subfigure}
\caption{
    Example likelihood computations on the InvFels tree $\tau_1$ for fidelities $t=\{x_1,y_1,x_2,y_2,w\}$.
    Edges labeled by the probability of substitution along that edge.
    In (a), we compute the product to obtain $\Pr(\siteSplitRV=\{2,3\},\ancestralSplitRV=\emptyset\mid \tau_1,t) = (1+x_1)(1-x_2)(1+y_1)(1-y_2)(1+w)/32$.
    In (b), the same process yields $\Pr(\siteSplitRV=\{2,3\},\ancestralSplitRV=\{1\}\mid \tau_1,t) = (1+x_1)(1-x_2)(1+y_1)(1-y_2)(1-w)/32$.
}
\label{fig:example_likelihoods}
\end{figure}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\hline
$\siteSplit_j$ & $\ancestralSplit_k$ & $32\cdot\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau_1, t)$\\
\hline
$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
&$\{1\}^*$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$             \\
&$\{2\}^*$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$           \\

@@

We begin with the branch length inconsistency.

\blInconsist*
